Understanding the Lomb-Scargle Periodogram\\ {\it (Draft Version \today)}

### 2017-02-22 PracticalLombScargle.tex
More tweaks

\begin{abstract}
This paper presents a conceptual introduction to the Lomb-Scargle periodogram,
and discusses important practical aspects of its use to detect and characterize
periodic signals.
Rather than a rigorous mathematical treatment, the goal of this paper is to
build intuition about what assumptions are implicit in the use of the
periodogram, and to help develop the intuition required to interpret the
periodogram correctly and use it effectively.
\end{abstract}

@@

\subsection{Outline}
The remainder of this paper is organized as follows:
{\bf\sect{continuous-fourier-transform}} presents a review of the
continous Fourier transform and some of its useful properties,
including defining the notion of a power spectrum (i.e. Schuster
periodogram) for detecting periodic content in a signal.
{\bf\sect{window-functions}} builds on these properties to explore how
different observation patterns (i.e. window functions) affect the observed
power spectrum, and discusses a conceptual understanding of the
Nyquist limiting frequency.
{\bf\sect{non-uniform-sampling}} considers non-uniformly sampled signals as
a special case of an observational window, and shows that the Nyquist-like
limit for this case is quite different than what many practitioners in
the field often assume.
{\bf\sect{schuster-to-lomb-scargle}} introduces the Lomb-Scargle periodogram,
a modified version of the classical periodogram for unevenly-sampled data,
as well as the motivation behind these modifications.
{\bf\sect{lomb-scargle-extensions}} discusses a complementary view of the
Lomb-Scargle periodogram as the result of least-squares model fitting,
as well as some extensions enabled by that viewpoint.
{\bf\sect{practical-considerations}} utilizes the concepts introduced
in earlier sections to discuss important practical considerations for
the use of the Lomb-Scargle periodogram, including the choice of
frequency grid, uncertainties and false alarm probabilities, and modes of
failure that should be accounted for when working with real-world data.
{\bf\sect{conclusion}} concludes and summarizes the key recommendations
for practical use of the Lomb-Scargle periodogram.

@@

Finally, there has been some exploration of extensions to the Lomb-Scargle
periodogram for use with multi-band observations, using various forms of
regularization to control model complexity \citep{VanderPlas2015, Long2016}.
With many of these least-squares and/or Bayesian extensions,
computational complexity quickly becomes an issue, because fast
$\mathcal{O}(N\log N)$ approaches which can be used for sinusoidal fits
(see \sect{algorithmic-considerations}) are not available for more general
functional forms, though there is some promising work in this area: see,
for example, the {\it Fast Template Periodogram} (Hoffman et al. 2017,
{\it in prep}) which can quickly construct periodograms from Fourier
approximations to templates.

@@

\subsection{Postscript: Why Lomb-Scargle?}
\sectlabel{postscript}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers lean toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.

As discussed in \sect{schuster-to-lomb-scargle}, the periodogram has two
benefits over the Schuster approach:
the noise distribution at each individual frequency is the sum
of two Gaussians, and the result is equivalent to a periodogram derived
from a least squares analysis.
But somehow along the way, a mythology seems to have developed surrounding
the efficiency and efficacy of the Lomb-Scargle approach. In particular,
it's common to see statements or implications along the lines of the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with little extra
    computation \citep[see the discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant issue of
    maximum peak heights across multiple frequencies,
    which must be either approximated
    or computed by bootstrap resampling (see \sect{false-alarm-probability})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions, the Lomb-Scargle
    periodogram has the same window-driven issues as the Schuster
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: Bayesian
    analysis shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding the frequency of a generic, potentially
    non-sinusoidal signal (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths set aside, what is the practical advantage of Lomb-Scargle
over a Schuster approach?
What would we lose if we instead used a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

@@

All computations and figures in this paper were produced using Python,
and in particular used the
Numpy \citep{numpy, numpybook},
Pandas \citep{pandas},
Astropy \citep{Astropy2013},
and Matplotlib \citep{matplotlib} packages.
Periodograms where computed using the
Astropy\footnote{\url{http://docs.astropy.org/en/stable/stats/lombscargle.html}}
implementations of the \citet{Press89}, \citet{Zechmeister09}, and
\citet{Palmer09} algorithms, which were adapted from Python code originally
published by \citet{ICVG2014} and \citet{VanderPlas2015}.
All code behind this paper, including code to reproduce all figures,
is available in the form of Jupyter notebooks in the paper's GitHub
repository\footnote{\url{http://github.com/jakevdp/PracticalLombScargle/}}.

\acknowledgments
{\it Acknowledgments:} I am grateful to Jeff Scargle, Max Mahlke,
Zeljko Ivezic, and Andy Connolly for helpful feedback on early drafts
of this paper.
This work was supported by the University of Washington eScience Institute,
with funding from the Gordon and Betty Moore Foundation, the Alfred P. Sloan
Foundation, and the Washington Research Foundation.

>>>>

\begin{abstract}
The Lomb-Scargle periodogram is a well-known algorithm for detecting and
characterizing periodic signals in unevenly-sampled data.
This paper presents a conceptual introduction to the Lomb-Scargle periodogram
and important practical considerations for its use.
Rather than a rigorous mathematical treatment, the goal of this paper is to
build intuition about what assumptions are implicit in the use of the
periodogram, so as to motivate important practical considerations required
in its application and interpretation.
\end{abstract}

@@

\subsection{Outline}
The remainder of this paper is organized as follows:\\
{\bf\sect{continuous-fourier-transform}} presents a review of the
continous Fourier transform and some of its useful properties,
including defining the notion of a power spectrum (i.e. Schuster
periodogram) for detecting periodic content in a signal.\\
{\bf\sect{window-functions}} builds on these properties to explore how
different observation patterns (i.e. window functions) affect the observed
power spectrum, and discusses a conceptual understanding of the
Nyquist limiting frequency.\\
{\bf\sect{non-uniform-sampling}} considers non-uniformly sampled signals as
a special case of an observational window, and shows that the Nyquist-like
limit for this case is quite different than what many practitioners in
the field often assume.\\
{\bf\sect{schuster-to-lomb-scargle}} introduces the Lomb-Scargle periodogram,
a modified version of the classical periodogram for unevenly-sampled data,
as well as the motivation behind these modifications.\\
{\bf\sect{lomb-scargle-extensions}} discusses a complementary view of the
Lomb-Scargle periodogram as the result of least-squares model fitting,
as well as some extensions enabled by that viewpoint.\\
{\bf\sect{practical-considerations}} builds on the concepts introduced
in earlier sections to discuss important practical considerations for
the use of the Lomb-Scargle periodogram, including the choice of
frequency grid, uncertainties and false alarm probabilities, and modes of
failure that should be accounted for when working with real-world data.\\
{\bf\sect{conclusion}} concludes and summarizes the key recommendations
for practical use of the Lomb-Scargle periodogram.

@@

Finally, there has been some exploration of extensions to the Lomb-Scargle
periodogram for use with multi-band observations, using various forms of
regularization to control model complexity \citep{VanderPlas2015, Long2016}.
With many of these least-squares and/or Bayesian extensions,
computational complexity quickly becomes an issue, because fast
$\mathcal{O}(N\log N)$ approaches which can be used for sinusoidal fits
(see \sect{algorithmic-considerations}) are not available for more general
functional forms, though there is some promising work in this area: see,
for example, the {\it Fast Template Periodogram} (Hoffman et al. 2017,
{\it in prep}\footnote{\url{http://ascl.net/code/v/1559}})
which can quickly construct periodograms from Fourier approximations to templates.

@@

\subsection{Postscript: Why Lomb-Scargle?}
\sectlabel{postscript}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers tend to gravitate toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.

As discussed in \sect{schuster-to-lomb-scargle}, the periodogram has two
benefits over the Schuster approach:
the noise distribution at each individual frequency is the sum
of two Gaussians, and the result is equivalent to a periodogram derived
from a least squares analysis.
But somehow along the way, a mythology seems to have developed surrounding
the efficiency and efficacy of the Lomb-Scargle approach. In particular,
it's common to see statements or implications along the lines of the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with little extra
    computation \citep[see the discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant question of
    maximum peak heights across multiple frequencies,
    which must be either approximated
    or computed by bootstrap resampling (see \sect{uncertainties})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions common to most astronomical
    applications, the Lomb-Scargle
    periodogram has the same window-driven issues as the Schuster
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: Bayesian
    analysis shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding the frequency of a generic, potentially
    non-sinusoidal signal (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths set aside, what is the practical advantage of Lomb-Scargle
over a Schuster approach?
What would we lose if we instead used a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

@@

All computations and figures in this paper were produced using Python,
and in particular used the
Numpy \citep{numpy, numpybook},
Pandas \citep{pandas},
AstroPy \citep{Astropy2013},
and Matplotlib \citep{matplotlib} packages.
Periodograms where computed using the
AstroPy implementations\footnote{The AstroPy Lomb-Scargle documentation is at
\url{http://docs.astropy.org/en/stable/stats/lombscargle.html}}
of the \citet{Press89}, \citet{Zechmeister09}, and
\citet{Palmer09} algorithms, which were adapted from Python code originally
published by \citet{ICVG2014} and \citet{VanderPlas2015}.
All code behind this paper, including code to reproduce all figures,
is available in the form of Jupyter notebooks in the paper's GitHub
repository\footnote{Source code \& notebooks available at
\url{http://github.com/jakevdp/PracticalLombScargle/}}.

\acknowledgments
{\it Acknowledgments:} I am grateful to Jeff Scargle, Max Mahlke,
and Zeljko Ivezic for helpful feedback on early drafts
of this paper.
This work was supported by the University of Washington eScience Institute,
with funding from the Gordon and Betty Moore Foundation, the Alfred P. Sloan
Foundation, and the Washington Research Foundation.
